---https://github.com/Xuuuyeah/p8130_final.git
title: "p8130_final"
author: "Yiran Xu"
date: "2024-12-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 目前遇到的问题：
建模用的是AFT model, assume "weibull" distribution. 但是输出结果后发现survival months不属于"weibull" distribution (根据残差是否是normal的方法),同样也拒绝了log- normal distribution.

由于找不到假设的分布，使用非参数方法COX model后cross validation的结果很差，mean_cindex只有0.2+，所以目前先保留了AFT的方法，还在看怎么解决不符合分布的问题。。。

### 剩下的任务：

* outlier detection: Cook's distance



* 可以在尝试一下LASSO，但这个dataset dimension小，不一定适合用lasso，stepwise就够
* evaluation: prediction risk = bias^2 + var?
* readability

### Yiran To-do list
* 还没有完成不同种族间的比较 - stratification and rerun Weibull
* model里只有三个main effect, 应该把其他的main effect都加上？（之前因为有interaction所以没加上）
* model explanation: need to output "survival_month = 0.1x + 0.2y + ..."
* month不需要transform，因为假设的不是normal
* If the interaction term is NOT significant, remove it and re-fit the model only with the main effects
* use the compound variable "Regional Node Positive/Regional Node Examined", instead of the two individual variables 


#### Done
* transformation, reginol_node_positive还没找到合适的transformation，survival
 - whatever
* 怎么解决不符合分布的问题 (已解决,未修改：according to lec 12: In large samples, the test is very sensitive to departures from normality, we could still assume a distribution given the residual is not normal, use qq plot is enough) 


```{r}
library(readr)
library(dplyr)
library(survival)
library(caret)
library(ggplot2)
library(patchwork)
library(kableExtra)
library(minerva)
library(stats)
library(fitdistrplus)
library(randomForestSRC)
library(tidyverse)
library(goftest)
library(survival)
library(stats)
library(broom)
```
# Data Wrangling

## Loading and cleaning the data

```{r}
data_clean = read_csv("/Users/boxiangtang/Desktop/Biosta_HW/Biosta_final/Project_2_data.csv") |>
  rename(
    Stage_6th = `6th Stage`) |>
  janitor::clean_names() |>
  
mutate(status = as.numeric(status == "Dead"),
       race = as.factor(race),
       marital_status = as.factor(marital_status),
       t_stage = as.factor(t_stage),
       n_stage = as.factor(n_stage),
       stage_6th = as.factor(stage_6th),
       differentiate = as.factor(differentiate),
       grade = as.factor(grade),
       a_stage = as.factor(a_stage),
       estrogen_status = as.factor(estrogen_status),
       progesterone_status = as.factor(progesterone_status),
       survival_months = log(survival_months) + 1
  )


# correct the col name: from "reginol_node_positive" to "regional_node_positive" (TBX)
names(data_clean)[names(data_clean) == "reginol_node_positive"] <- "regional_node_positive"


summary(data_clean)

head(data_clean) |>
  kable() |>
  kable_styling(full_width = FALSE) %>%
  scroll_box(width = "100%", height = "300px")
```

## Exploratory Data Analysis
```{r}
summary_stats <- data_clean |> summary()
print(summary_stats)

data_clean |> ggplot(aes(x = survival_months, fill = status)) +
  geom_histogram() +
  labs(title = "Survival Time Distribution", x = "Months", y = "Count") +
  theme_minimal()
```
Create a table for the description of each variable

```{r}
# Create a data frame for variable descriptions
variable_descriptions <- data.frame(
  Variable = c(
    "Age", "Race", "Marital Status", "T Stage (Tumor)", "N Stage (Node)",
    "Stage (6th Edition)", "Differentiate", "Grade", "A Stage", 
    "Tumor Size", "Estrogen Status", "Progesterone Status",
    "Regional Nodes Examined", "Regional Nodes Positive", 
    "Survival Months", "Status"
  ),
  Description = c(
    "Patient's age at the time of diagnosis or study enrollment.",
    "Patient's racial identity: Black, White, Other.",
    "Patient's marital status: Married, Single, Divorced, Separated, Widowed.",
    "Tumor size and extent: T1 (≤2 cm), T2 (>2 cm but ≤5 cm), T3 (>5 cm), T4 (invasion into chest wall or skin).",
    "Lymph node involvement: N1 (1–3 nodes), N2 (4–9 nodes), N3 (≥10 nodes).",
    "Overall cancer stage: IIA, IIB, IIIA, IIIB, IIIC.",
    "Tumor differentiation level: Well, Moderately, Poorly, Undifferentiated.",
    "Tumor histological grade: Grade 1 (low), Grade 2 (moderate), Grade 3 (high), Grade IV (anaplastic).",
    "Extent of cancer spread: Regional (local spread), Distant (metastasized).",
    "Size of the tumor in millimeters.",
    "Tumor's estrogen receptor status: Positive, Negative.",
    "Tumor's progesterone receptor status: Positive, Negative.",
    "Number of regional lymph nodes examined for cancer.",
    "Number of regional lymph nodes found to be cancer-positive.",
    "Number of months the patient survived after diagnosis or study enrollment.",
    "Patient's status at the end of the study: Alive, Deceased."
  )
)

# Display the table
kable(variable_descriptions, col.names = c("Variable", "Description"), 
      caption = "Variable Descriptions in the Breast Cancer Dataset", align = "l")
```



Side-by-Side Histograms for Categorical Variables
```{r}
# Histogram for `race`
plot_race <- ggplot(data_clean, aes(x = race)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(title = "Race", x = "Race", y = "Frequency") +
  theme_minimal(base_size = 12)

# Histogram for `marital_status`
plot_marital_status <- ggplot(data_clean, aes(x = marital_status)) +
  geom_bar(fill = "lightgreen", color = "black") +
  labs(title = "Marital Status", x = "Marital Status", y = "Frequency") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Histogram for `t_stage`
plot_t_stage <- ggplot(data_clean, aes(x = t_stage)) +
  geom_bar(fill = "salmon", color = "black") +
  labs(title = "T Stage", x = "T Stage", y = "Frequency") +
  theme_minimal(base_size = 12)

# Histogram for `n_stage`
plot_n_stage <- ggplot(data_clean, aes(x = n_stage)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(title = "N Stage", x = "N Stage", y = "Frequency") +
  theme_minimal(base_size = 12)

# Histogram for `stage_6th`
plot_stage_6th <- ggplot(data_clean, aes(x = stage_6th)) +
  geom_bar(fill = "gold", color = "black") +
  labs(title = "Stage (6th Edition)", x = "Stage (6th Edition)", y = "Frequency") +
  theme_minimal(base_size = 12)

# Histogram for `differentiate`
plot_differentiate <- ggplot(data_clean, aes(x = differentiate)) +
  geom_bar(fill = "orchid", color = "black") +
  labs(title = "Differentiate", x = "Differentiate", y = "Frequency") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Histogram for `grade`
plot_grade <- ggplot(data_clean, aes(x = as.factor(grade))) +
  geom_bar(fill = "coral", color = "black") +
  labs(title = "Grade", x = "Grade", y = "Frequency") +
  theme_minimal(base_size = 12)

# Histogram for `a_stage`
plot_a_stage <- ggplot(data_clean, aes(x = a_stage)) +
  geom_bar(fill = "plum", color = "black") +
  labs(title = "A Stage", x = "A Stage", y = "Frequency") +
  theme_minimal(base_size = 12)

# Histogram for `estrogen_status`
plot_estrogen_status <- ggplot(data_clean, aes(x = estrogen_status)) +
  geom_bar(fill = "lightblue", color = "black") +
  labs(title = "Estrogen Status", x = "Estrogen Status", y = "Frequency") +
  theme_minimal(base_size = 12)

# Histogram for `progesterone_status`
plot_progesterone_status <- ggplot(data_clean, aes(x = progesterone_status)) +
  geom_bar(fill = "yellow", color = "black") +
  labs(title = "Progesterone Status", x = "Progesterone Status", y = "Frequency") +
  theme_minimal(base_size = 12)

# Histogram for `status`
plot_status <- ggplot(data_clean, aes(x = as.factor(status))) +
  geom_bar(fill = "lightgreen", color = "black") +
  labs(title = "Status", x = "Status", y = "Frequency") +
  theme_minimal(base_size = 12)

# Combine all plots into a grid using patchwork
## Panel 1: First set of plots
panel_1 <- (plot_race | plot_marital_status | plot_t_stage) /
           (plot_n_stage | plot_stage_6th | plot_differentiate)

## Panel 2: Second set of plots
panel_2 <- (plot_grade | plot_a_stage | plot_estrogen_status) /
           (plot_progesterone_status | plot_status)

## Display Panel 1
panel_1

## Display Panel 2
panel_2

```


# Pie Chart for Categorical Variables (TBX)
```{r}

# Ensure variables are converted to factor type
convert_to_factor <- function(data, vars) {
  data %>% mutate(across(all_of(vars), as.factor))
}

# Convert variables to factor type
factor_vars <- c("race", "marital_status", "t_stage", "n_stage", 
                 "stage_6th", "differentiate", "grade", "a_stage", 
                 "estrogen_status", "progesterone_status", "status")
data_clean <- convert_to_factor(data_clean, factor_vars)

# Define a function to simplify long labels
simplify_labels <- function(label, max_length = 12) {
  ifelse(nchar(label) > max_length, paste0(substr(label, 1, max_length), "..."), label)
}

# Define a function to plot pie charts
plot_pie <- function(data, var, title) {
  counts <- data %>% count(!!sym(var)) %>% mutate(percent = round(n / sum(n) * 100, 1))
  counts <- counts %>% mutate(label = paste0(simplify_labels(as.character(!!sym(var))), " (", percent, "%)"))
  pie_data <- counts$n
  labels <- counts$label
  pie(pie_data, labels = labels, main = title, col = rainbow(length(pie_data)))
}

# Display in multiple pages
# Page 1
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot_pie(data_clean, "race", "Race")
plot_pie(data_clean, "marital_status", "Marital Status")
plot_pie(data_clean, "t_stage", "T Stage")
plot_pie(data_clean, "n_stage", "N Stage")

# Page 2
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot_pie(data_clean, "stage_6th", "Stage (6th Edition)")
plot_pie(data_clean, "differentiate", "Differentiate")
plot_pie(data_clean, "grade", "Grade")
plot_pie(data_clean, "a_stage", "A Stage")

# Page 3
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot_pie(data_clean, "estrogen_status", "Estrogen Status")
plot_pie(data_clean, "progesterone_status", "Progesterone Status")
plot_pie(data_clean, "status", "Status")

```

Side-by-Side Histograms for Numerical Variables
```{r}
## Create Individual Plots for Numeric Variables
plot_age <- ggplot(data_clean, aes(x = age)) +
  geom_histogram(fill = "steelblue", color = "black", bins = 10) +
  labs(title = "Age", x = "Age", y = "Frequency") +
  theme_minimal(base_size = 12)

plot_tumor_size <- ggplot(data_clean, aes(x = tumor_size)) +
  geom_histogram(fill = "salmon", color = "black", bins = 10) +
  labs(title = "Tumor Size", x = "Tumor Size", y = "Frequency") +
  theme_minimal(base_size = 12)

plot_regional_node_examined <- ggplot(data_clean, aes(x = regional_node_examined)) +
  geom_histogram(fill = "lightgreen", color = "black", bins = 10) +
  labs(title = "Regional Nodes Examined", x = "Regional Nodes Examined", y = "Frequency") +
  theme_minimal(base_size = 12)

plot_regional_node_positive <- ggplot(data_clean, aes(x = regional_node_positive)) +
  geom_histogram(fill = "plum", color = "black", bins = 10) +
  labs(title = "Regional Nodes Positive", x = "Regional Nodes Positive", y = "Frequency") +
  theme_minimal(base_size = 12)

plot_survival_months <- ggplot(data_clean, aes(x = survival_months)) +
  geom_histogram(fill = "gold", color = "black", bins = 10) +
  labs(title = "Survival Months", x = "Survival Months", y = "Frequency") +
  theme_minimal(base_size = 12)

(plot_age | plot_tumor_size | plot_regional_node_examined) /
                 (plot_regional_node_positive | plot_survival_months)
```


# Box Plot for Numerical Variables (TBX)
```{r}
# Select numeric variables and remove missing values
num_vars <- c("age", "tumor_size", "regional_node_examined", "regional_node_positive", "survival_months")

# Use base R to filter the required columns
data_clean_numeric <- data_clean[, num_vars, drop = FALSE]

# Remove rows with missing values
data_clean_numeric <- na.omit(data_clean_numeric)

# Plot boxplots for each numeric variable
par(mfrow = c(2, 3)) # Set layout to 2 rows and 3 columns

boxplot(data_clean_numeric$age,
        main = "Age",
        ylab = "Age",
        col = "steelblue",
        border = "black")

boxplot(data_clean_numeric$tumor_size,
        main = "Tumor Size",
        ylab = "Tumor Size",
        col = "salmon",
        border = "black")

boxplot(data_clean_numeric$regional_node_examined,
        main = "Regional Nodes Examined",
        ylab = "Regional Nodes Examined",
        col = "lightgreen",
        border = "black")

boxplot(data_clean_numeric$regional_node_positive,
        main = "Regional Nodes Positive",
        ylab = "Regional Nodes Positive",
        col = "plum",
        border = "black")

boxplot(data_clean_numeric$survival_months,
        main = "Survival Months",
        ylab = "Survival Months",
        col = "gold",
        border = "black")

# Reset plot parameters to default
par(mfrow = c(1, 1))

```



# Transformation
```{r}
data_clean$log_tumor_size <- log1p(data_clean$tumor_size) 
data_clean$log_nodes_examined <- log1p(data_clean$regional_node_examined)
# data_clean$log_nodes_positive <- log1p(data_clean$regional_node_positive)
# not useful
# data_clean$log_survival_months <- log1p(data_clean$survival_months)
# not useful


## Create individual plots
plot_log_tumor_size <- ggplot(data_clean, aes(x = log_tumor_size)) +
  geom_histogram(fill = "steelblue", color = "black", bins = 10) +
  labs(title = "Log Tumor Size", x = "Log(Tumor Size)", y = "Frequency") +
  theme_minimal(base_size = 12)

plot_log_nodes_examined <- ggplot(data_clean, aes(x = log_nodes_examined)) +
  geom_histogram(fill = "salmon", color = "black", bins = 10) +
  labs(title = "Log Regional Nodes Examined", x = "Log(Regional Nodes Examined)", y = "Frequency") +
  theme_minimal(base_size = 12)


  theme_minimal(base_size = 12)

## Combine the plots into a panel using patchwork
(plot_log_tumor_size | plot_log_nodes_examined) 
```



# Identify the influential outliers (TBX)
```{r}
# Step 1: Define variables for analysis
variables_to_check <- c("tumor_size", "regional_node_examined", "regional_node_positive", "survival_months")

# Step 2: Function to calculate Cook's Distance for a single variable
check_influential_points <- function(data, variable) {
  # Fit a simple linear regression model with the selected variable as predictor
  formula <- as.formula(paste("survival_months ~", variable))
  model <- lm(formula, data = data)
  
  # Calculate Cook's Distance
  cooks_dist <- cooks.distance(model)
  
  # Identify outliers using boxplot stats
  outliers <- which(data[[variable]] %in% boxplot.stats(data[[variable]])$out)
  
  # Identify influential points based on Cook's Distance threshold
  threshold <- 4 / nrow(data)
  influential_points <- which(cooks_dist > threshold)
  
  # Cross-check influential points with outliers
  influential_outliers <- intersect(outliers, influential_points)
  
  # Summarize results
  summary_table <- data.frame(
    Row_Index = outliers,
    Variable = variable,
    Cook_Distance = cooks_dist[outliers],
    Influential = ifelse(outliers %in% influential_outliers, "Yes", "No")
  )
  
  # Add decision column
  summary_table <- summary_table %>%
    mutate(Decision = ifelse(Influential == "Yes", "Consider Removal", "Retain"))
  
  return(summary_table)
}

# Step 3: Apply the function to all variables and combine results
all_results <- lapply(variables_to_check, function(var) check_influential_points(data_clean, var))
combined_results <- bind_rows(all_results)

# Step 4: Display the head of the combined summary table
print("Head of Outliers and Influential Points Summary:")
print(head(combined_results))

# Step 5: Summary of Results
total_outliers <- nrow(combined_results)
influential_outliers <- sum(combined_results$Decision == "Consider Removal")

cat("\nSummary of Results:\n")
cat("Total Outliers Identified:", total_outliers, "\n")
cat("Outliers Identified as Influential Points:", influential_outliers, "\n")

if (influential_outliers > 0) {
  cat("Some outliers are influential and should be reviewed for potential removal.\n")
} else {
  cat("No influential outliers detected. All points may be retained.\n")
}


```

# Compare the models with & without influential outliers (TBX)
```{r}
# Model with influential points
model_with <- lm(survival_months ~ ., data = data_clean)

# Model without influential points
data_clean_no_influential <- data_clean[-combined_results$Row_Index[combined_results$Decision == "Consider Removal"], ]
model_without <- lm(survival_months ~ ., data = data_clean_no_influential)

# Compare coefficients
summary(model_with)
summary(model_without)
```

# Check the multi-collinearity between numerical variables
```{r}
# Load required libraries
library(graphics)

# Select numerical variables (adjust to your dataset)
numerical_vars <- data_clean[, c("tumor_size", "regional_node_examined", "regional_node_positive", "survival_months")]

# Create a scatterplot matrix
pairs(
  numerical_vars,
  panel = function(x, y) {
    points(x, y, pch = 20, col = "blue", cex = 0.5)  # Add scatterplot points
    abline(lm(y ~ x), col = "red", lwd = 2)          # Add red trend line
  },
  main = "Scatterplot Matrix with Trend Lines"
)
```



# 接下来想要决定是否删除那些identify出来的influential points，但是我还没有想好具体的实施方法，然后很多类似check的方法或者像box-cox这种寻找最优transformation的方法都需要先fit一个lm model，但我觉得咱们目前没有fit lm model的任何步骤。
```{r}
# 提取系数
coeff_with <- coef(model_with)
coeff_without <- coef(model_without)

# 计算变化百分比
coeff_change <- (coeff_without - coeff_with) / coeff_with * 100

# 创建表格
coeff_table <- data.frame(
  Variable = names(coeff_with),           # 变量名称
  Coeff_With_Influential = coeff_with,    # 包含影响点的系数
  Coeff_Without_Influential = coeff_without, # 删除影响点后的系数
  Percent_Change = coeff_change           # 百分比变化
)

# 查看表格
print(coeff_table)

# 如果需要更美观的表格输出，使用 knitr 或 gt
library(knitr)
kable(coeff_table, digits = 2, caption = "Coefficient Changes with and without Influential Points")


```






## optional: 数值变量分布： 可以绘制直方图或箱线图检查是否有极端值?

# Apply AFT method

## Check survival rate distribution.
```{r}
# Weibull 分布拟合
fit_weibull <- fitdist(data_clean$survival_months, "weibull")

ad_test <- ad.test(data_clean$survival_months, "pweibull", 
                   shape = fit_weibull$estimate["shape"], 
                   scale = fit_weibull$estimate["scale"])
print("Anderson-Darling Test:")
print(ad_test)

fit_lognormal <- fitdist(data_clean$survival_months, "lnorm")
plot(fit_lognormal)


# 绘制拟合效果
plot(fit_weibull)
```

The figure above shows that the distribution is either not Weibull or log-normal test, this is because the survival months data is truncated. If the residual is normal distribution, then we can say the theoriotical survival months obey certain distribution.


## Stepwise: initial filtering of variables

survival rate distribution is closed to Weibull distribution --> adopt it as the hypothetical distribution.

### stepwise
```{r}
full_model <- survreg(Surv(survival_months, status) ~ ., data = data_clean, dist = "weibull")


stepwise_model <- stepAIC(full_model, direction = "both", trace = FALSE) 

selected_formula <- formula(stepwise_model)
selected_variables <- all.vars(selected_formula)[-1][-1]
print(selected_variables)
```

`6th stage`,`Differentiation`,`A stage`,`Tumor Size` are excluded, as are explained by T Stage alone?


```{r}
selected_vars <- all.vars(selected_formula)[-1][-1]
selected_vars = data_clean[, selected_vars, drop = FALSE]

perform_chi_square_all <- function(data) {
  results <- data.frame(
    Variable1 = character(),
    Variable2 = character(),
    Chi_Square_Statistic = numeric(),
    P_Value = numeric(),
    stringsAsFactors = FALSE
  )
  
  # 遍历所有变量组合
  var_names <- names(data)
  for (i in 1:(length(var_names) - 1)) {
    for (j in (i + 1):length(var_names)) {
      var1 <- var_names[i]
      var2 <- var_names[j]
      
      # 卡方检验
      contingency_table <- table(data[[var1]], data[[var2]])
      chi_test <- chisq.test(contingency_table)
      
      # 存储结果
      results <- rbind(results, data.frame(
        Variable1 = var1,
        Variable2 = var2,
        Chi_Square_Statistic = chi_test$statistic,
        P_Value = chi_test$p.value
      ))
    }
  }
  significant_results <- subset(results, P_Value < 0.05)
  return(significant_results)
}

# 假设 data_clean 全是因子类型数据
# 应用到你的数据集
chi_square_results <- perform_chi_square_all(selected_vars) |>
  arrange(P_Value)

# 打印结果
print(chi_square_results)

```

Select the top ten interaction from ANOVA result and added them to the model

Only selcted `age`, `race` and `marital_status` as main effect. unlike other corelated pathological factor, these factors are either less likely to correlated with each other or correlated with pathological factors.
```{r}
top_interactions <- head(chi_square_results, 10)
print("Top 10 Interactions:")
print(top_interactions)

# 构建交互项公式
interaction_terms <- paste0(top_interactions$Variable1, " * ", top_interactions$Variable2)
interaction_formula <- paste(interaction_terms, collapse = " + ")
main_terms = paste(c("age", "race"), collapse = " + ")

# 构建最终 AFT 模型公式
aft_formula <- as.formula(paste("Surv(survival_months, status) ~ ", paste(c(interaction_formula, main_terms), collapse = "+")))
print("Final AFT Formula:")
print(aft_formula)
```

# Model output
```{r}
aft_model_output <- survreg(aft_formula, data = data_clean, dist = "weibull")

# Tidy the model output
tidy_fit <- tidy(aft_model_output, conf.int = TRUE)

# Add exponentiated coefficients for time ratios
tidy_fit <- tidy_fit %>%
  mutate(
    exp_coefficient = exp(estimate),
    exp_conf_low = exp(conf.low),
    exp_conf_high = exp(conf.high)
  ) %>%
  dplyr::select(term, estimate, exp_coefficient, exp_conf_low, exp_conf_high, p.value )

# Print the readable table
tidy_fit %>%
  print(n = Inf)
```

# Evaluate the model with 5-fold cross validation
# 5-fold cross validation
```{r}
cross_validate_aft_with_residuals <- function(data, formula, dist = "weibull", k = 5) {
  cindex_results <- c()
  residuals_list <- list()
  set.seed(123)  # 确保结果可复现
  folds <- cut(seq(1, nrow(data)), breaks = k, labels = FALSE)
  
  for (i in 1:k) {
    # 划分训练集和测试集
    test_idx <- which(folds == i, arr.ind = TRUE)
    train_data <- data[-test_idx, ]
    test_data <- data[test_idx, ]
    
    # 拟合 AFT 模型
    aft_model <- survreg(formula, data = train_data, dist = dist)
    
    # 预测测试集生存时间
    predictions <- predict(aft_model, newdata = test_data, type = "response")
    
    # 计算 C-index
    concordance_result <- concordance(Surv(test_data$survival_months, test_data$status) ~ predictions)
    cindex_results <- c(cindex_results, concordance_result$concordance)
    
    # 计算残差
    train_residuals <- residuals(aft_model, type = "response")
    residuals_list[[i]] <- train_residuals
    
    
    # 绘制残差分布图（仅示例第一个 fold）
    if (i == 1) {
      print("Residuals Distribution for Fold 1:")
      hist(train_residuals, main = "Residuals Distribution", xlab = "Residuals")
      qqnorm(train_residuals, main = "QQ Plot of Residuals")
      qqline(train_residuals, col = "red")
    }
  }
  
  # 返回交叉验证结果和残差信息
  return(list(
    mean_cindex = mean(cindex_results),
    sd_cindex = sd(cindex_results),
    residuals = residuals_list
  ))
}

# 执行交叉验证并检查残差
cv_results_with_residuals <- cross_validate_aft_with_residuals(data_clean, aft_formula, dist = "weibull", k = 5)


# 输出交叉验证结果
print("5-Fold Cross-Validation Results with Residual Analysis:")
print(cv_results_with_residuals$mean_cindex)
print(cv_results_with_residuals$sd_cindex)
```

# Outliers
```{r}
aft_model_outliner <- survreg(aft_formula, data = data_clean, dist = "weibull")

# Residuals
residuals <- residuals(aft_model_outliner, type = "deviance")

# Leverage (hat values)
X <- model.matrix(aft_model_outliner)  # Design matrix
H <- X %*% solve(t(X) %*% X) %*% t(X)  # Hat matrix
leverage <- diag(H)

# Number of parameters
p <- length(coef(aft_model_outliner))

# Cook's Distance
cooks_distance <- (residuals^2 / p) * (leverage / (1 - leverage)^2)

# Plot Cook's distance
plot(cooks_distance, type = "h", main = "Cook's Distance", xlab = "Observation", ylab = "Cook's Distance")
abline(h = 4 / nrow(data_clean), col = "red", lty = 2)

# Identify observations with high Cook's distance
outliers <- which(cooks_distance > 4 / nrow(data_clean))
print(unname(outliers))
```

# Reference

